---
title: "Conditional Panel Segments in **lattice** `stripplot`"
date: "A long time ago in a galaxy far, far away..."
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background

I came across an interesting question with the topic of [“Condition layer across panels in lattice”](http://stackoverflow.com/questions/35353923/condition-layer-across-panels-in-lattice/35476469#35476469) asked recently on StackOverflow. The questioner was trying to add line segments from an auxiliary dataset to a **lattice** `stripplot` using `layer` from **latticeExtra**, but failed since the group-specific vertical lines were displayed in each of the sub-panels, i.e. regardless of the particular group.

It made me think about how I would tackle this issue and I finally came up with two different solutions on how to prevent the conditional line segments from being drawn in each of the sub-panels. But first of all, here is the sample data.

```{r}
## load packages
library(latticeExtra)

## create sample data
raw_data = data.frame(subject = rep(1:6, 4),
                       cond1 = as.factor(rep(1:2, each = 12)),
                       cond2 = rep(rep(c("A", "B"), each = 6), 2),
                       response = c(2:7, 6:11, 3:8, 7:12))

## create auxiliary data
summary_data =
  data.frame(cond1 = as.factor(rep(1:2, each = 2)),
             cond2 = rep(c("A", "B"), times = 2),
             mean = aggregate(response ~ cond2 * cond1, raw_data, mean)$response,
             within_ci = c(0.57, 0.54, 0.6, 0.63))

summary_data$lci = summary_data$mean - summary_data$within_ci
summary_data$uci = summary_data$mean + summary_data$within_ci

## create stripplot
p_strip = stripplot(response ~ cond1 | cond2, groups = subject,
                     data = raw_data, panel = function(x, y, ...) {
                       panel.stripplot(x, y, type = "b", lty = 2, ...)
                       panel.average(x, y, fun = mean, lwd = 2,
                                     col = "black", ...)
                     })

print(p_strip)

```


## Approach #1: manual segment insertion via trellis.focus

Now, my first solution was to print the stripplot (e.g., inside a png or any other graphics device) and subsequently modify each sub-panel using `trellis.focus`.

```{r}
## display stripplot
print(p_strip)

## loop over grops
for (i in c("A", "B")) {

  # subset of current group
  dat = subset(summary_data, cond2 == i)

  # add intervals to current panel
  trellis.focus(name = "panel", column = ifelse(i == "A", 1, 2), row = 1)
  panel.segments(x0 = dat$cond1, y0 = dat$lci,
                 x1 = dat$cond1, y1 = dat$uci, subscripts = TRUE)
  trellis.unfocus()
}

```


## Approach #2: conditional segments via panel.number

Although this approach is quite straightforward, it has a big disadvantage: the created plot cannot be stores in a variable which excludes its use for subsequent processing inside R. Therefore, I came up with another (possibly more convenient) solution and created a separate xyplot where the group-specific lower and upper y-values (y0, y1 passed on to panel.segments) were set manually in dependence of the current panel.number. The two separate plots were subsequently blended together using as.layer from **latticeExtra**. In contrast to the first approach using `trellis.focus`, the thus created plot can be stored in a variable and is hence available for subsequent processing inside R.


```{r}
p_seg = xyplot(lci ~ cond1 | cond2, data = summary_data, ylim = c(1, 13),
       panel = function(...) {
         # lower and upper y values
         y0 = list(summary_data$lci[c(1, 3)], summary_data$lci[c(2, 4)])
         y1 = list(summary_data$uci[c(1, 3)], summary_data$uci[c(2, 4)])
         # insert vertical lines depending on current panel
         panel.segments(x0 = 1:2, x1 = 1:2,
                        y0 = y0[[panel.number()]],
                        y1 = y1[[panel.number()]])
       })

p_comb = p_strip +
  as.layer(p_seg)

print(p_comb)

```

